---
title: "Is inflation increasing faster than Wage growth? Is gender wage gap real?"
author: "Sagarika Ramesh"
description: "Exploring the USA Bureau of Labor Statistics (BLS) data to see if inflation is increasing faster than wage growth and if the gender wage gap is real."
output:
  distill::distill_article:
    self_contained: false
---

I want to explore the USA Bureau of Labor Statistics (BLS) data. I will be analyzing different trends and patterns that exist in the data. Is inflation increasing faster than wage growth? If it is true, it would be difficult for lower-income households to meet their daily needs.

My hypothesis is inflation is increasing faster than wage growth. The prices of most of the products/services have increased. We always hear people complain about how their spending power has decreased significantly. We can now check if it is true. 
Dataset is obtained from  https://www.bls.gov/

My sample consists of following variables:  
1. Consumer Price Index for All Urban Consumers(CPI): All Items in U.S. City Average. It is a measure of the average change over time in the prices paid by urban consumers for a market basket of consumer goods and services.    
2. Employment Cost Index(ECI). It essentially measures the change in total employee compensation each quarter.  
3. Difference between CPI growth and ECI growth. It measures the difference between CPI growth and ECI growth. CPI change measures the increase or decrease in the index from the previous year. ECI change measures the increase or decrease in the index from the previous year.  
4. Employment Rate.  
5. Government Current Expenditures. It consists of spending by government to produce and provide services to the public, such as national defense and public school education.  
6. Personal consumption expenditures (PCE). It is the primary measure of consumer spending on goods and services in the U.S. economy.  
7. Corporate Profits After Tax. It represents the portion of the total income earned from current
production that is accounted for by U.S. corporations.

My outcome variable/Dependent variable: Difference between CPI change and ECI change. 
I have constructed a hypothesis test to check my hypothesis. Research design: Before-and-after design. As I compare the outcomes each year, take the difference between the outcomes. 
Further, I will use regression analysis to examine the relationship between my dependent variable and independent variables. 

I am interested to test another hypothesis: Is gender wage gap real? In other words, do men earn more wages than women? I will be analyzing this as it is important to address inequality. 
Dataset is obtained from  https://www.bls.gov/ 

My sample consists of following variables:  
1. Median weekly earnings of Men and Women.    
2. Employment Rate of Men and Women.  
3. Family Households with Male/Female Householder.  

My outcome/dependent variable is Median weekly earnings of Men and Women. I have constructed a hypothesis test to check my hypothesis. Research design: Before-and-after design. As I compare the outcomes each year, take the difference between the outcomes.  
Further, I will use regression analysis to examine the relationship between my dependent variable and independent variables.

```{r}
library(blscrapeR)
library(dplyr)
library(tidyverse)
library(infer)
library(broom)
library(lubridate)
library(ggplot2)
```

```{r}
cpi_all_items = read.csv("data/CPIAUCSL.csv")

cpi_all_items$Date <- ymd(cpi_all_items$DATE) 
cpi_all_items <- cpi_all_items|>
            rename("cpi_all_items" = CPIAUCSL)|>
            select(Date, cpi_all_items)

cpi_all_items$year <- year(cpi_all_items$Date)
cpi_all_items$month <- month(cpi_all_items$Date)

cpi_all_items <- cpi_all_items|>
  filter(month == 01 | month == 4 | month == 7| month == 10)|>
  select(Date, cpi_all_items)

cpi_all_items$CPI_change <- (cpi_all_items$cpi_all_items - lag(cpi_all_items$cpi_all_items))

glimpse(cpi_all_items)
head(cpi_all_items)
```

CPI_change measures the increase or decrease in the index from the previous year

```{r}
eci_all  <- read.csv('data/ECIWAG.csv')

eci_all$Date <- ymd(eci_all$DATE) 
eci_all <- eci_all|>
            rename("eci" = ECIWAG)|>
            select(Date, eci)

eci_all$eci_change <- eci_all$eci - lag(eci_all$eci)

glimpse(eci_all)
head(eci_all)
```

eci_change measures the increase or decrease in the index from the previous year. 

```{r}
cpi_eci <- cpi_all_items |>
          inner_join(eci_all, by = 'Date')|>
          select(Date, CPI_change, eci_change)

cpi_eci$diff_btw_cpi_eci <- cpi_eci$CPI_change - cpi_eci$eci_change

head(cpi_eci)
```
diff_btw_cpi_eci is the difference between CPI_change and eci_change. It measures the difference between CPI growth and ECI growth. 

```{r}
ggplot(data = cpi_eci, mapping = aes(y = CPI_change, x = eci_change)) + geom_point(color = "purple") + geom_smooth(method = "loess")+
  labs(
  y = "Change in CPI",
  x = "Change in ECI",
  title = "Change in ECI vs Change in CPI") + 
  scale_x_continuous( limits = c(0, 7),
                 breaks = c(0, 1,2,3,4,5,6,7)) +
  scale_y_continuous(limits = c(0, 7),
                 breaks = c(0, 1,2,3,4,5,6,7))
```


```{r}
ggplot(cpi_eci, aes(x = Date, y = diff_btw_cpi_eci)) + geom_line() + labs(x = "Date", y = "Difference between CPI and ECI", title = "Difference between CPI and ECI over a period of 10 years")
```
The independent variables(employment rate, Government Current Expenditures, Personal Consumption Expenditures, Corporate Profits After Tax) are added to the same data set.

```{r}
employment_rate <- read.csv('data/LREM64TTUSM156S.csv')

employment_rate$Date <- ymd(employment_rate$DATE) 
employment_rate <- employment_rate|>
            rename("emp_rate" = LREM64TTUSM156S)|>
            select(Date, emp_rate)

glimpse(employment_rate)
head(employment_rate)
```

```{r}
gov_cur_expend <- read.csv('data/GEXPND.csv')

gov_cur_expend$Date <- ymd(gov_cur_expend$DATE) 
gov_cur_expend <- gov_cur_expend|>
            rename("gov_cur_expend" = GEXPND)|>
            select(Date, gov_cur_expend)

glimpse(gov_cur_expend)
head(gov_cur_expend)
```

```{r}
personal_consumption_expend <- read.csv('data/PCE.csv')

personal_consumption_expend$Date <- ymd(personal_consumption_expend$DATE) 
personal_consumption_expend <- personal_consumption_expend|>
            rename("pce" = PCE)|>
            select(Date, pce)

glimpse(personal_consumption_expend)
head(personal_consumption_expend)
```

```{r}
corp_profit_after_tax <- read.csv('data/CPATAX.csv')

corp_profit_after_tax$Date <- ymd(corp_profit_after_tax$DATE) 
corp_profit_after_tax <- corp_profit_after_tax|>
            rename("corp_profit" = CPATAX)|>
            select(Date, corp_profit)

glimpse(corp_profit_after_tax)
head(corp_profit_after_tax)
```

```{r}
finance_data <- cpi_eci|>
                left_join(gov_cur_expend, by = 'Date')

finance_data <- finance_data|>
                left_join(employment_rate, by = 'Date')

finance_data <- finance_data|>
                left_join(corp_profit_after_tax, by = 'Date')

finance_data <- finance_data|>
               left_join(personal_consumption_expend, by = 'Date')|>
                select(!(CPI_change:eci_change))



glimpse(finance_data)
head(finance_data)
```

```{r}
ggplot(data = finance_data, aes(x = emp_rate, y = diff_btw_cpi_eci)) + geom_point(color = "purple") +labs(x = "Employment Rate", y = "Difference between CPI and ECI growth", title = " Employment Rate vs Difference between CPI and ECI growth")
```
```{r}
ggplot(data = finance_data, aes(x = gov_cur_expend, y = diff_btw_cpi_eci)) + geom_point(color = "purple")+labs(x = "Government Current Expenditures", y = "Difference between CPI and ECI growth", title = "Government Current Expenditures vs Difference between CPI and ECI growth")
```
```{r}
ggplot(data = finance_data, aes(x = pce, y = diff_btw_cpi_eci)) + geom_point(color = "purple") + labs(x = "Personal Consumption Expenditures", y = "Difference between CPI and ECI growth", title = "Personal Consumption Expenditures vs Difference between CPI and ECI growth")
```
```{r}
ggplot(data = finance_data, aes(x = corp_profit, y = diff_btw_cpi_eci)) + geom_point(color = "purple") + labs(x = "Corporate Profits after Tax", y = "Difference between CPI and ECI growth", title = "Corporate Profits vs Difference between CPI and ECI growth")
```


```{r}
observed_mean = mean(finance_data$diff_btw_cpi_eci, na.rm = TRUE)
observed_mean
```
On average the CPI increases by 0.66 points more than ECI.  

Hypothesis Testing:  
Null Hypothesis: Average difference between CPI growth and ECI growth is 0. Mean(diff_btw_cpi_eci) is 0.  
Alternate Hypothesis: Average difference between CPI growth and ECI growth is more than 0. Mean(diff_btw_cpi_eci) is more than 0.  


```{r}
set.seed(7890)

null_dist <- finance_data |>
              specify(response = diff_btw_cpi_eci) |>
              hypothesize(null = "point", mu = 0) |>
              generate(reps = 1000, type = "bootstrap") |>
              calculate(stat = "mean")
null_dist
```

```{r}
visualize(null_dist) + 
  shade_p_value(obs_stat = observed_mean, direction = "right")

```

```{r}
null_dist|> 
  get_p_value(obs_stat = observed_mean, direction = "right")
```
The probability of observing a difference between CPI growth and ECI growth as large as 0.66 points due to sampling variation alone in the null distribution is 0.005 = 0.5%. 
Since this p_value is smaller than our significance level of 0.05. We reject the null hypothesis. We have enough evidence to say CPI growth is faster than ECI growth. In other words, inflation is increasing faster than Wage growth. 

Multiple regression: 
```{r}
finance_data.multreg <- lm(diff_btw_cpi_eci ~  emp_rate + gov_cur_expend + pce + corp_profit, data = finance_data) 

finance_data.multreg
tidy(finance_data.multreg) |> knitr::kable(digits = 3)
```

```{r}
modelsummary::modelsummary(finance_data.multreg, statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"), gof_map = c("nobs", "r.squared", "adj.r.squared"))
```
The coefficient of emp_rate is -0.663, which means there is a decrease in the Difference between CPI growth and ECI growth(outcome) by 0.663 points with a 1% increase in employment rate, when other variables don’t change.

The coefficient of gov_cur_expend is -0.002, which means there is a decrease in the Difference between CPI growth and ECI growth(outcome) by 0.002 points with a one-unit(1 billion dollars) increase in Government Current Expenditures, when other variables don’t change.

The coefficient of pce is 0.002, which means there is an increase in the Difference between CPI growth and ECI growth(outcome) by 0.002 points with a one-unit(1 billion dollars) increase in Personal Consumption Expenditures, when other variables don’t change.

The coefficient of corp_profit is 0.002, which means there is an increase in the Difference between CPI growth and ECI growth(outcome) by 0.002 points with a one-unit(1 billion dollars) increase in Corporate Profits, when other variables don’t change.

The estimated coefficient of emp_rate is statistically significant at the 0.05 level because the p-value 0.006, which is less than 0.05.
The estimated coefficient of pce is statistically significant at the 0.05 level because the p-value is less than 0.05.  
The estimated coefficient of gov_cur_expend	 is statistically significant at the 0.05 level because the p-value is less than 0.05.
The estimated coefficient of corp_profit is not statistically significant at the 0.05 level because the p-value 0.381, which is more than 0.05.


Now, lets start another interesting hypothesis.  
Is gender wage gap real?
```{r}
totalwage_weekly <- read.csv("data/LES1252881500Q.csv")

totalwage_weekly$Date <- ymd(totalwage_weekly$DATE) 
totalwage_weekly <- totalwage_weekly|>
            rename("tt_wk_wage" = LES1252881500Q)|>
            select(Date, tt_wk_wage)

glimpse(totalwage_weekly)
head(totalwage_weekly)
```
```{r}
ggplot(totalwage_weekly, aes(x = Date, y = tt_wk_wage)) + geom_line() + labs(x = "Date", y = "Dollars",title = "Median weekly earnings over 10 years for ages 16 and above")
```

```{r}
malewage_weekly <- read.csv("data/LES1252881800Q.csv")

malewage_weekly$Date <- ymd(malewage_weekly$DATE) 
malewage_weekly <- malewage_weekly|>
            rename("male_wk_wage" = LES1252881800Q)|>
            select(Date, male_wk_wage)

glimpse(malewage_weekly)
head(malewage_weekly)
```
```{r}
ggplot(malewage_weekly, aes(x = Date, y = male_wk_wage)) + geom_line()+ labs(x = "Date", y = "Dollars", title = "Median weekly earnings over 10 years for male of ages 16 and above")
```

```{r}
femalewage_weekly <- read.csv("data/LES1252882700Q.csv")

femalewage_weekly$Date <- ymd(femalewage_weekly$DATE) 
femalewage_weekly <- femalewage_weekly|>
            rename("female_wk_wage" = LES1252882700Q)|>
            select(Date, female_wk_wage)

glimpse(femalewage_weekly)
head(femalewage_weekly)
```

```{r}
ggplot(femalewage_weekly, aes(x = Date, y = female_wk_wage)) + geom_line() + labs(x = "Date", y = "Dollars", title = "Median weekly earnings over 10 years for female of ages 16 and above")
```
```{r}
wages <- totalwage_weekly |>
  inner_join(malewage_weekly, by = 'Date')

wages <- wages |>
        inner_join(femalewage_weekly, by = 'Date')
glimpse(wages)
head(wages)
```

```{r}
ggplot(wages, aes(x = male_wk_wage, y = female_wk_wage)) +
geom_point(color = "steelblue1") +
labs(
  x = "Median weekly earnings by Male",
  y = "Median weekly earnings by Female",
  title = "Median weekly earnings by Male vs Female") + 
  scale_x_continuous(labels = scales::dollar, 
                        limits = c(690, 1200),
                 breaks = c(600, 700, 800, 900, 1000, 1100, 1200)) +
  scale_y_continuous(labels = scales::dollar, 
                     limits = c(690, 1200),
                 breaks = c(600, 700, 800, 900, 1000, 1100, 1200))
```

```{r}
wages <- wages|>
        pivot_longer(
          cols = male_wk_wage:female_wk_wage,
          names_to = "Gender",
    values_to = "Wage"
  )|> mutate(Gender = if_else(Gender == "male_wk_wage", "Male", "Female"))|>
  select(-tt_wk_wage)

head(wages)
```


```{r}
emp_rate_female <- read.csv('data/LREM25FEUSM156S.csv')

emp_rate_female$Date <- ymd(emp_rate_female$DATE) 
emp_rate_female <- emp_rate_female|>
            rename("emp_rate_female" = LREM25FEUSM156S)|>
            select(Date, emp_rate_female)

glimpse(emp_rate_female)
head(emp_rate_female)
```

```{r}
emp_rate_male <- read.csv('data/LREM25MAUSM156S.csv')

emp_rate_male$Date <- ymd(emp_rate_male$DATE) 
emp_rate_male <- emp_rate_male|>
            rename("emp_rate_male" = LREM25MAUSM156S)|>
            select(Date, emp_rate_male)

glimpse(emp_rate_male)
head(emp_rate_male)
```
```{r}
emp_rate <- emp_rate_male|>
            inner_join(emp_rate_female, by='Date')

emp_rate <- emp_rate|>
        pivot_longer(
          cols = emp_rate_male:emp_rate_female,
          names_to = "Gender",
    values_to = "emp_rate"
  )|> mutate(Gender = if_else(Gender == "emp_rate_male", "Male", "Female"))

head(emp_rate)
```


```{r}
female_householder <- read.csv('data/OFHHFH.csv')

female_householder$Date <- ymd(female_householder$DATE) 
female_householder <- female_householder|>
            rename("female_householder" = OFHHFH)|>
            select(Date, female_householder)

glimpse(female_householder)
head(female_householder)

```


```{r}
male_householder <- read.csv('data/OFHHMH.csv')

male_householder$Date <- ymd(male_householder$DATE) 
male_householder <- male_householder|>
            rename("male_householder" = OFHHMH)|>
            select(Date, male_householder)

glimpse(male_householder)
head(male_householder)
```

```{r}
house_holder <- male_householder|>
            inner_join(female_householder, by='Date')


house_holder <- house_holder|>
        pivot_longer(
          cols = male_householder:female_householder,
          names_to = "Gender",
    values_to = "house_holder"
  )|> mutate(Gender = if_else(Gender == "male_householder", "Male", "Female"))

head(house_holder)
```
```{r}
wages <- left_join(wages,emp_rate, by = c('Date'='Date', 'Gender'='Gender'))

wages <-left_join(wages,house_holder, by = c('Date'='Date', 'Gender'='Gender'))

glimpse(wages)
head(wages)
```
```{r}
ggplot(data = wages, mapping = aes(x=Gender, y =emp_rate)) +
geom_boxplot(na.rm = TRUE)+ labs(x = "Gender", y = "Employment Rare", title = "Gender vs Employment Rate")

```
```{r}
ggplot(data = wages, mapping = aes(x=Gender, y =Wage)) +
geom_boxplot(na.rm = TRUE)+ labs(x = "Gender", y = "Wage", title = "Gender vs Wage")

```


```{r}
est_ate <-  wages |>
       specify(Wage ~ Gender) |>
      calculate(stat = "diff in means", order = c("Male", "Female"))
est_ate
```
On average, weekly earning by male is 174.33 dollars more than Female weekly earning.

Hypothesis Testing:  
Null Hypothesis: Wages of Male and Female are equal.  
Alternate Hypothesis: Wages of Male is more than Female.

```{r}
null_dist_wage_diff <- wages |>
                    specify(Wage ~ Gender) |>
                    hypothesize(null = "independence") |>
                    generate(reps = 1000, type = "permute") |>
                    calculate(stat = "diff in means", order = c("Male", "Female"))

ate_pval <- null_dist_wage_diff |>
get_p_value(obs_stat = est_ate, direction = "right")
ate_pval
```


```{r}
null_dist_wage_diff |>
visualize() +
shade_p_value(obs_stat = est_ate, direction = "right")
```
The probability of observing a difference between male and female weekly wages as large as 174.33 dollars due to sampling variation alone in the null distribution is close to 0.
Since this p_value is smaller than our significance level of 0.05. We reject the null hypothesis. We have enough evidence to say male weekly wakes is more than female weekly wages.

```{r}
wages.multreg <- lm(Wage ~  Gender + emp_rate + house_holder, data = wages) 

wages.multreg
tidy(wages.multreg) |> knitr::kable(digits = 3)
```

```{r}
modelsummary::modelsummary(wages.multreg, statistic = c("s.e. = {std.error}",
                                         "p = {p.value}"),gof_map = c("nobs", "r.squared", "adj.r.squared"))

```
Baseline for comparison is female worker. 
The coefficient of emp_rate is 35.242, which means there is an increase in the weekly wages by 35.242 dollars with a 1% increase in the employment rate, when other variables don’t change.

The coefficient of house_holder is 0.190, which means there is an increase in the weekly wages by 0.190 dollars with a one-unit increase in Household(thousand households), when other variables don’t change.

The estimated coefficient of emp_rate is statistically significant at the 0.05 level because the p-value is less than 0.05.
The estimated coefficient of house_holder is statistically significant at the 0.05 level because the p-value is less than 0.05.  
The estimated coefficient of GenderMale	 is statistically significant at the 0.05 level because the p-value is less than 0.05.


Finally, I was able to successfully run two hypothesis tests. Two Multiple linear regression tests were done to find which independent variable is statistically significant. There can be other confounding variables which I might have not added to my data set. The results can be improved by adding more independent variables.  
